\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-13 k-planar.tex
Last undefined reference
Small updates

  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{compactenum}
Observe that, for any edge $vw\in E(G)$, 
\begin{equation}  
  \dist_{G^+}(v,w) \le k+1 \enspace , 
\end{equation}
since (the curve) $vw$ contains a sequence $W_{vw}$ of at most $k+2$ vertices of $G^+$ that form a walk from $v$ to $w$ in $G^+$.

@@

Our proof of \thmref{1-planar} follows quickly from the following technical lemma, which is an extension of the equivalent result for plane graphs \cite{dujmovic.joret.ea:planar}.
\begin{lem}\lemlabel{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS tree of $G'$ rooted at some vertex $r$.
    \item For each integer $j\ge 0$, let $V_j=\{v\in V(G):\dist_T(r,v)=j\}$. 
    \item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
    \begin{compactenum} 
      \item No edge of $F$ is crossed by any edge of $G$; and
      \item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $F[P_i]$ is a path; and
        \item $|V(P_i)\cap V_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N^+$ and $N^-$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\{S_x : x\in V(H)\}$ such that
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap V_j|\le 15$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

@@

  The remainder of the proof finishes in the same way as the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $V(Y^+-F)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.

  The planarity of $H$ comes from two properties:
  \begin{enumerate}
    \item $G/\mathcal{P}$ and $G'\mathcal{P}$ are isomorphic.  This follows immediately from the fact that, for each edge $xy\in E(G-G')$, $x,y\in S_a$ for some $a\in V(H)$.  Indeed, $S_a=Y^+_a$ where $\overline{Y}_a$ contains the edge $vw\in E(G')$ that crosses $xy$.

    \item $G[\overline{Y}^+-F]$ is connected. To see why this is so, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y^+}$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
  \end{enumerate}
  Since $G'$ is planar, this implies that $H=G'/\mathcal{P}$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}
\end{proof}

>>>>

  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{compactenum}
Observe that, for any edge $vw\in E(G)$, 
\begin{equation}  
  \dist_{G^+}(v,w) \le k+1 \enspace ,  \eqlabel{distance-preserving}
\end{equation}
since (the curve) $vw$ contains a sequence $W_{vw}$ of at most $k+2$ vertices of $G^+$ that form a walk from $v$ to $w$ in $G^+$.

@@

Our proof of \thmref{1-planar} follows quickly from the following technical lemma, which is an extension of the equivalent result for plane graphs \cite{dujmovic.joret.ea:planar}.
\begin{lem}\lemlabel{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS tree of $G'$ rooted at some vertex $r$.
    \item For each integer $j\ge 0$, let $V_j=\{v\in V(G):\dist_T(r,v)=j\}$. 
    \item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
    \begin{compactenum} 
      \item No edge of $F$ is crossed by any edge of $G$; and
      \item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $F[P_i]$ is a path; and
        \item $|V(P_i)\cap V_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N^+$ and $N^-$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\mathcal{P}=\{S_x : x\in V(H)\}$ such that
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap V_j|\le 15$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

@@

  The remainder of the proof finishes in the same way as the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $V(Y^+)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.

  The planarity of $H$ comes from two properties:
  \begin{enumerate}
    \item $G/\mathcal{P}$ and $G'/\mathcal{P}$ are isomorphic.  This follows immediately from the fact that, for each edge $xy\in E(G-G')$, $x,y\in S_a$ for some $a\in V(H)$.  Indeed, $S_a=Y^+_a$ where $\overline{Y}_a$ contains the edge $vw\in E(G')$ that crosses $xy$.

    \item $G[\overline{Y}^+-F]$ is connected. To see why this is so, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y}^+$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
  \end{enumerate}
  Since $G'$ is planar, this implies that $H=G'/\mathcal{P}$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}
\end{proof}
